\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-15 papers/jss/article.Rnw
rewrite of section 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@788 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Modern data collection and analysis pipelines involve collections
of components to enhance conrol of complex systems through
reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for various phases of data analysis -- collection, 
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
each environment or language. Every stage of such a data
analysis pipeline may produce intermediate results that need to be
stored or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?

@@

Traditionally, scientists and statisticians often use character seperated
text formats such as \texttt{CSV} \citep{shafranovich2005common} to 
export and import data. However, anyone who has ever used this will have
noticed that this method has many limitations: it is restricted to tabular 
datasets, lacks type-safety, and has limited precision for numeric values.
Moreover, ambiguities in the format itself frequently cause problems. 
For example the default characters used as seperator and decimal point 
are different in various parts of the world.
\emph{Extensible Markup Language} (\texttt{XML}) is another text-based 
well-established and widely-supported format with the ability to define
just about any arbitrarily complex schema \citep{nolan2013xml}. However,
it pays for this complexity with comparatively large and verbose messages,
and added complexities at the parsing side (which are somewhat mitigated
by the availability of mature libraries and parsers).
A more modern, widely used format is \emph{JavaScript Object Notation}
(\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}. This format is text-based as well and used mostly
on the web. Several R packages implement functions to parse and generate
\texttt{JSON} data from R objects. A number of \texttt{JSON} dialects has
been proposed, such as \texttt{BSON} and \texttt{MessagePack} which both
add binary support. However, these derivatives are not compatible with 
existing JSON software, and have not been widely adopted.

@@

- This paper introduces another format: protocol buffers
- unique combination of features that make it very suitable for numerical computing:
- binary, schema, versioned, mature, high quality cross language implementations
- we argue that (complex) statistical applications will benefit from using this format

%we should probably explain what a schema is%
Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. Many sources compare data serialization formats
and show Protocol Buffers compare very favorably to the alternatives; see
\citet{Sumaray:2012:CDS:2184751.2184810} for one such comparison.
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages

@@

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers can be described as a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  Since their
introduction, Protocol Buffers have been widely adopted in industry with
applications as varied as %database-internal messaging (Drizzle), % DE: citation?
Sony Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  
% TODO(DE): This either needs a citation, or remove the name drop
% MS: These are mostly from blog posts, I can't find a good reference
% that has a long list, and the name and year citation style seems
% less conducive to long lists of marginal citations like blog posts
% compared to say concise CS/math style citations [3,4,5,6]. Thoughts?
While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is compartively flexible and simple to use.

>>>>

Modern data collection and analysis pipelines involve collections
of decoupled components in order to manage and control complexity 
through reusability, modularity, and fault isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
% This is really a different pattern not connected well here.
%Data analysis patterns such as Split-Apply-Combine
%\citep{wickham2011split} explicitly break up large problems into manageable pieces.  
These pipelines are frequently built using different programming 
languages for various phases of data analysis -- collection, 
cleaning, modeling, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
each environment or language. Every stage of such a data
analysis pipeline may produce intermediate results that need to be
stored or sent over the network for further processing. 
% JO Perhaps also mention that serialization is needed for distributed
% systems to make systems scale up?

@@

Traditionally, scientists and statisticians often use character seperated
text formats such as \texttt{CSV} \citep{shafranovich2005common} to 
export and import data. However, anyone who has ever used \texttt{CSV} will have
noticed that this method has many limitations: it is restricted to tabular 
data, lacks type-safety, and has limited precision for numeric values.
Moreover, ambiguities in the format itself frequently cause problems. 
For example, conventions on which characters used as seperator and decimal
point vary by country.
\emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define
just about any arbitrarily complex schema \citep{nolan2013xml}. However,
it pays for this complexity with comparatively large and verbose messages,
and added complexitiy at the parsing side (which are somewhat mitigated
by the availability of mature libraries and parsers). Because \texttt{XML}
is text based and has no native notion of numeric types or arrays, it 
usually not a very practical format to store numeric datasets as they appear
in statistical applications.
A more modern, widely used format is \emph{JavaScript Object Notation}
(\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used mostly on the web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings, 
booleans and null. However, because it is a text-based format, numbers are
stored as human-readable decimal notation which is somewhat inefficient and
leads to loss of type (double vs integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects. 
A number of \texttt{JSON} dialects has been proposed, such as \texttt{BSON} and 
\texttt{MessagePack} which both add binary support. However, these derivatives
are not compatible with existing JSON software, and have not been widely adopted.

@@

In 2008, Google released an open source version of Protocol Buffers: the data 
interchange format that was designed and used for their internal infrastructure.
Google officially provides high quality parsing libraries for \texttt{Java}, 
\texttt{C++} and \texttt{Python}, and community developed open source implementations
are available for many other languages. 
Protocol Buffers take a quite different approach from many other popular formats.
They offer a unique combination of features, performance, and maturity that seems
particulary well suited for data-driven applications and numerical computing.
Protocol Buffers are a binary format that natively supports all common primitive types
found in modern programming languages. The advantage of this is that numeric values
are serialized exactly the same way as they are stored in memory. Therefore there is
no loss of precision, no overhead, and parsing messages is very efficient: the system can 
simply copy bytes to memory without any further processing. 
But the most powerful feature of protocol buffers is that it decouples the content
from the structure using a schema, very similar to a database. This further increases
performance by eliminating redundancy, while at the same time providing foundations
for defining an \emph{Interface Description Language}, or \emph{IDL}.
Many sources compare data serialization formats and show Protocol Buffers compare 
very favorably to the alternatives; see \citet{Sumaray:2012:CDS:2184751.2184810} 
for one such comparison.

@@

% JO: I'm not sure where to put this paragraph. I think it is too technical
% for the introduction section. Maybe start this section with some explanation
% of what a schema is and then continue with showing how PB implement this?
Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. 
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers can be described as a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  Since their
introduction, Protocol Buffers have been widely adopted in industry with
applications as varied as %database-internal messaging (Drizzle), % DE: citation?
Sony Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  
% TODO(DE): This either needs a citation, or remove the name drop
% MS: These are mostly from blog posts, I can't find a good reference
% that has a long list, and the name and year citation style seems
% less conducive to long lists of marginal citations like blog posts
% compared to say concise CS/math style citations [3,4,5,6]. Thoughts?

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is compartively flexible and simple to use.
